---
  title: "STAT 405 Course Project"
subtitle: "Exploratory Data Analysis and Bayesian Modeling of Emotional Dynamics"
author:
  - name: Yu Chang
email: changy72@student.ubc.ca
- name: Marc Shi
email: lianjies@student.ubc.ca
date: today
format:
  pdf:
  toc: true
number-sections: true
colorlinks: true
fig-pos: "H"
geometry: margin=1in
fontsize: 11pt
execute:
  echo: true
warning: false
message: false
editor: visual
---

Predictive Power:

```{r}

library(readr)
library(dplyr)
library(tidyr)
library(purrr)

# ---------------------------
# 1) Read data
# ---------------------------
dat <- read_tsv("Data/Backup/backup.tsv", show_col_types = FALSE)

emotions <- c(
  "angry", "enthusiastic", "happy", "irritated",
  "nervous", "pleased", "relaxed", "sad", "stressed"
)

# ---------------------------
# 2) Build lag-1 transitions
#    - sort within subject
#    - keep only truly adjacent prompts
#    - within-subject center emotions
# ---------------------------
prepare_lagged_data <- function(df, emotions, beeps_per_day = 5) {
  df %>%
    arrange(id, day, beep) %>%
    group_by(id) %>%
    mutate(
      slot = (day - 1L) * beeps_per_day + beep
    ) %>%
    mutate(
      across(
        all_of(emotions),
        ~ .x - mean(.x, na.rm = TRUE),
        .names = "{.col}_c"
      )
    ) %>%
    mutate(
      next_slot = lead(slot)
    ) %>%
    mutate(
      across(
        all_of(c(emotions, paste0(emotions, "_c"))),
        lead,
        .names = "next_{.col}"
      )
    ) %>%
    ungroup() %>%
    filter(next_slot == slot + 1L)
}

lagged_dat <- prepare_lagged_data(dat, emotions)

# ---------------------------
# 3) Score one predictor-target pair
#    Goal:
#      predict target_(t+1)
#      baseline:   target_(t+1) ~ target_t
#      augmented:  target_(t+1) ~ target_t + predictor_t
#
#    CV:
#      leave-one-subject-out
# ---------------------------
score_one_pair <- function(data, predictor, target) {
  pred_c <- paste0(predictor, "_c")
  targ_c <- paste0(target, "_c")
  next_targ_c <- paste0("next_", target, "_c")

  dd <- data %>%
    select(id, all_of(c(pred_c, targ_c, next_targ_c))) %>%
    drop_na()

  ids <- sort(unique(dd$id))

  fold_scores <- map_dfr(ids, function(test_id) {
    train <- dd %>% filter(id != test_id)
    test  <- dd %>% filter(id == test_id)

    if (nrow(train) < 20 || nrow(test) == 0) {
      return(NULL)
    }

    fit0 <- lm(
      formula = reformulate(termlabels = targ_c, response = next_targ_c),
      data = train
    )

    fit1 <- lm(
      formula = reformulate(termlabels = c(targ_c, pred_c), response = next_targ_c),
      data = train
    )

    p0 <- predict(fit0, newdata = test)
    p1 <- predict(fit1, newdata = test)
    y  <- test[[next_targ_c]]

    tibble(
      id = test_id,
      n_test = nrow(test),
      sse0 = sum((y - p0)^2),
      sse1 = sum((y - p1)^2)
    )
  })

  if (nrow(fold_scores) == 0) {
    return(tibble(
      predictor = predictor,
      target = target,
      rmse_baseline = NA_real_,
      rmse_augmented = NA_real_,
      rmse_gain = NA_real_,
      rel_mse_reduction = NA_real_,
      n_obs = 0L
    ))
  }

  mse0 <- sum(fold_scores$sse0) / sum(fold_scores$n_test)
  mse1 <- sum(fold_scores$sse1) / sum(fold_scores$n_test)

  tibble(
    predictor = predictor,
    target = target,
    rmse_baseline = sqrt(mse0),
    rmse_augmented = sqrt(mse1),
    rmse_gain = sqrt(mse0) - sqrt(mse1),             # bigger = better
    rel_mse_reduction = (mse0 - mse1) / mse0,        # bigger = better
    n_obs = sum(fold_scores$n_test)
  )
}

# ---------------------------
# 4) Run all predictor-target pairs
# ---------------------------
pair_grid <- expand_grid(
  predictor = emotions,
  target = emotions
) %>%
  filter(predictor != target)

pair_scores <- pmap_dfr(
  pair_grid,
  ~ score_one_pair(lagged_dat, ..1, ..2)
)

# ---------------------------
# 5) Overall ranking of emotions
# ---------------------------
overall_scores <- pair_scores %>%
  group_by(predictor) %>%
  summarise(
    total_rmse_gain = sum(rmse_gain, na.rm = TRUE),
    mean_rmse_gain = mean(rmse_gain, na.rm = TRUE),
    mean_rel_mse_reduction = mean(rel_mse_reduction, na.rm = TRUE),
    median_rel_mse_reduction = median(rel_mse_reduction, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_rel_mse_reduction), desc(total_rmse_gain))

overall_scores
pair_scores %>% arrange(desc(rel_mse_reduction))

# ---------------------------
# 6) Top 3 targets helped by each predictor
# ---------------------------
top_targets_per_predictor <- pair_scores %>%
  group_by(predictor) %>%
  arrange(desc(rel_mse_reduction), .by_group = TRUE) %>%
  slice_head(n = 3) %>%
  ungroup()

top_targets_per_predictor

# 按照real_mse_reduction的大小排序，然后每个predictor选三个

#Interpretation of rel_mse_reduction for the pair: happy - pleased: 在控制 pleased 自己上一时刻的值以后，加入 happy_t 可以让对 pleased_(t+1) 的样本外预测误差下降大约 1.55%。
# ---------------------------
# 7) Winner
# ---------------------------
overall_scores

# total_rmse_gain: 把这个predictor对所有target的rmse_gain全加起来
#mean_rel_mse_reeduction: 这个predictor对所有其他targets的平均相对MSE改善比例

ggsave("~/Desktop/overall.pdf", plot = p_corr,
       width = 8, height = 6)
```

Spearman correlation

```{r}

#| label: corrplot-all-emotions
#| fig-width: 7
#| fig-height: 6
#| message: false
#| warning: false

library(tidyverse)

# select all emotion variables
emotion_vars <- c(
  "angry", "enthusiastic", "happy", "irritated",
  "nervous", "pleased", "relaxed", "sad", "stressed"
)

emotion_data <- backup %>%
  select(all_of(emotion_vars))

# Spearman correlation matrix
corr_mat <- cor(
  emotion_data,
  use = "pairwise.complete.obs",
  method = "spearman"
)

# turn into long format for ggplot
corr_df <- as.data.frame(as.table(corr_mat))

p_corr <- ggplot(corr_df, aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", Freq)), size = 3) +
  scale_fill_gradient2(low = "steelblue4", mid = "white", high = "firebrick3",
                       midpoint = 0, limits = c(-1, 1)) +
  theme_minimal() +
  labs(
    title = "Spearman Correlation Plot for All Emotion Variables",
    x = NULL,
    y = NULL,
    fill = "Correlation"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )

p_corr
ggsave("~/Desktop/correlation_heatmap.pdf", plot = p_corr,
       width = 8, height = 6)
```

## ACF

## Autocorrelation Within Emotions

We first examine whether each emotion shows serial dependence over time. If repeated emotion measurements were approximately independent, then the autocorrelation function (ACF) should remain close to zero at small lags. In contrast, positive autocorrelation at nearby lags would suggest persistence in emotional intensity across adjacent prompts, which is more consistent with a sequential latent-state model than with an i.i.d. baseline.

```{r}

# libraries
library(tidyverse)
# data loading
backup <- readr::read_tsv("Data/Backup/backup.tsv", show_col_types = FALSE)

names(backup)
dim(backup)
head(backup)

# create average emotion series over day x beep
avg_series <- backup %>%
  group_by(day, beep) %>%
  summarise(
    happy = mean(happy, na.rm = TRUE),
    relaxed = mean(relaxed, na.rm = TRUE),
    sad = mean(sad, na.rm = TRUE),
    stressed = mean(stressed, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(day, beep)

avg_series


```

```{r}
avg_series$time_index <- 1:nrow(avg_series)

# time series plots
average_happy <- plot(avg_series$time_index, avg_series$happy, type = "l",
     main = "Average Happy Over Ordered Day x Beep",
     xlab = "Ordered day x beep index", ylab = "Mean happy")

pdf("average_happy_plot.pdf", width = 8, height = 5)

plot(avg_series$time_index, avg_series$happy, type = "l",
     main = "Average happy over Ordered Day x Beep",
     xlab = "ordered day x beep index", ylab = "Mean happy")

dev.off()

```

# plot for an unique participants on Happy

```{r}

# choose one random participant
random_id <- sample(unique(backup$id), 1)

# create ordered time index within participant
random_participant <- backup %>%
  filter(id == random_id) %>%
  arrange(day, beep) %>%
  mutate(time_index = row_number())

# plot happy time series for that participant
plot(random_participant$time_index,
     random_participant$happy,
     type = "l",
     lwd = 2,
     main = paste("Happy for Random Participant", random_id),
     xlab = "Ordered day x beep index",
     ylab = "Happy")
```

# ACF + CCF

```{r}
library(tidyverse)

# data loading

candidate_predictors
paste0(candidate_predictors, "_c")
setdiff(paste0(candidate_predictors, "_c"), names(lagged_dat))

backup <- readr::read_tsv("data/Backup/backup.tsv", show_col_types = FALSE)

# create average emotion series over ordered day x beep
avg_series <- backup %>%
  group_by(day, beep) %>%
  summarise(
    happy = mean(happy, na.rm = TRUE),
    pleased = mean(pleased, na.rm = TRUE),
    enthusiastic = mean(enthusiastic, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(day, beep)

# output one pdf with 3 plots in one row
pdf("happy_acf_ccf_combined.pdf", width = 12, height = 4)

par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

acf(
  avg_series$happy,
  main = "ACF: Happy"
)

ccf(
  avg_series$happy,
  avg_series$pleased,
  main = "CCF: Happy vs Pleased"
)

ccf(
  avg_series$happy,
  avg_series$enthusiastic,
  main = "CCF: Happy vs Enthusiastic"
)

dev.off()

# also show in RStudio
par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))

acf(
  avg_series$happy,
  main = "ACF: Happy"
)

ccf(
  avg_series$happy,
  avg_series$pleased,
  main = "CCF: Happy vs Pleased"
)

ccf(
  avg_series$happy,
  avg_series$enthusiastic,
  main = "CCF: Happy vs Enthusiastic"
)

par(mfrow = c(1, 1))


```

# Forward Selection

```{r}

library(tidyverse)

score_candidate_for_happy <- function(data, selected = character(0), candidate) {
  # ---------- clean selected ----------
  if (is.null(selected) || length(selected) == 0) {
    selected_terms <- character(0)
  } else {
    selected <- as.character(selected)
    selected <- selected[!is.na(selected)]
    selected <- selected[nzchar(selected)]
    selected_terms <- paste0(selected, "_c")
  }

  # ---------- clean candidate ----------
  candidate <- as.character(candidate)
  candidate <- candidate[!is.na(candidate)]
  candidate <- candidate[nzchar(candidate)]

  if (length(candidate) != 1) {
    stop("candidate must be exactly one non-empty string.")
  }

  candidate_term <- paste0(candidate, "_c")

  # ---------- formulas ----------
  base_terms <- c("happy_c", selected_terms)
  aug_terms  <- c(base_terms, candidate_term)

  needed_cols <- unique(c("id", "next_happy_c", aug_terms))

  # ---------- check columns ----------
  missing_cols <- setdiff(needed_cols, names(data))
  if (length(missing_cols) > 0) {
    stop("Missing columns in data: ", paste(missing_cols, collapse = ", "))
  }

  dd <- data %>%
    select(all_of(needed_cols)) %>%
    drop_na()

  ids <- sort(unique(dd$id))

  fold_scores <- purrr::map_dfr(ids, function(test_id) {
    train <- dd %>% filter(id != test_id)
    test  <- dd %>% filter(id == test_id)

    if (nrow(train) < 20 || nrow(test) == 0) {
      return(NULL)
    }

    fit_base <- lm(
      reformulate(termlabels = base_terms, response = "next_happy_c"),
      data = train
    )

    fit_aug <- lm(
      reformulate(termlabels = aug_terms, response = "next_happy_c"),
      data = train
    )

    p_base <- predict(fit_base, newdata = test)
    p_aug  <- predict(fit_aug, newdata = test)
    y_test <- test$next_happy_c

    tibble(
      id = test_id,
      n_test = nrow(test),
      sse_base = sum((y_test - p_base)^2),
      sse_aug  = sum((y_test - p_aug)^2)
    )
  })

  if (nrow(fold_scores) == 0) {
    return(tibble(
      candidate = candidate,
      rmse_base = NA_real_,
      rmse_aug = NA_real_,
      rmse_gain = NA_real_,
      rel_mse_reduction = NA_real_,
      n_obs = 0L
    ))
  }

  mse_base <- sum(fold_scores$sse_base) / sum(fold_scores$n_test)
  mse_aug  <- sum(fold_scores$sse_aug)  / sum(fold_scores$n_test)

  tibble(
    candidate = candidate,
    rmse_base = sqrt(mse_base),
    rmse_aug = sqrt(mse_aug),
    rmse_gain = sqrt(mse_base) - sqrt(mse_aug),
    rel_mse_reduction = (mse_base - mse_aug) / mse_base,
    n_obs = sum(fold_scores$n_test)
  )
}

forward_select_happy_predictors <- function(data, candidates, max_steps = 3) {
  candidates <- as.character(candidates)
  candidates <- candidates[!is.na(candidates)]
  candidates <- candidates[nzchar(candidates)]
  candidates <- unique(candidates)

  selected <- character(0)
  history <- list()

  for (step in seq_len(max_steps)) {
    remaining <- setdiff(candidates, selected)

    if (length(remaining) == 0) break

    step_scores <- purrr::map_dfr(
      remaining,
      function(x) score_candidate_for_happy(data, selected, x)
    ) %>%
      arrange(desc(rel_mse_reduction), desc(rmse_gain))

    best <- step_scores %>% slice(1)

    selected <- c(selected, best$candidate)

    history[[step]] <- list(
      step = step,
      selected_so_far = selected,
      ranking = step_scores,
      best = best
    )
  }

  list(
    selected = selected,
    history = history
  )
}

fs_result <- forward_select_happy_predictors(
  data = lagged_dat,
  candidates = candidate_predictors,
  max_steps = 3
)

fs_result$selected

forward_summary <- bind_rows(
  lapply(fs_result$history, function(x) {
    x$best %>%
      mutate(
        step = x$step,
        selected_so_far = paste(x$selected_so_far, collapse = ", ")
      ) %>%
      select(
        step, candidate, rmse_base, rmse_aug,
        rmse_gain, rel_mse_reduction, n_obs, selected_so_far
      )
  })
)

forward_summary

library(gridExtra)
library(grid)

pdf("forward_summary_table.pdf", width = 12, height = 3)

grid.table(forward_summary)

dev.off()

```
