---
title: "Model Fitting and Posterior Comparison"
author: "Marc Shi"
format: pdf
editor: visual
---

```{r}
library(cmdstanr)
library(readr)
library(dplyr)
set.seed(405)
```

# Data Importing & Cleaning

```{r}

dat <- read_tsv("Data/Backup/backup.tsv", show_col_types = FALSE)
```

from the dataset, we see that there are 7350 observations, which comes from 105 participants Ă— 14 days $\times 5$ beeps per day $=7350$. Therefore for each participants, we have $14 \times 5=70$ timepoints per participant.

```{r}
head(dat)
```

# Model Fitting

## 1. Naive Model (Baseline)

For the naive model, we use the **independent latent variable model with ordered logistic likelihood.**

## 1.1 Model set up

For each observation $i$ :

-   $y_i \in\{1,2,3,4,5\}$ : observed happiness rating (ordinal)

-   $x_i \in \mathbb{R}$ : latent continuous happiness

**Prior:** we assume $x_i \sim \operatorname{iid} \mathcal{N}\left(\mu, \sigma^2\right)$, where each latent happiness level is independent across time (within-participant) and between participants and is drawn from a same distribution with parameter. Since there is only one pbserved score $y_i$ connected to $x_i$ we rom the prior, we will be hierarchical and apply the hyper parameters as below:

$\mu \sim \mathcal{N}\left(0,1.5^2\right)$ and $\sigma \sim$ Exponential $(1)$

**Likelihood (Ordered Logistic Model)**

$$
y_i \mid x_i \sim \operatorname{Ordered} \operatorname{Logistic}\left(x_i, c\right)
$$

where: . $c=\left(c_1, c_2, c_3, c_4, c_5, c_6\right)$ are cutpoints with

$c = (-2.5,-1.5,-0.5,0.5,1.5,2.5)$, thus make $0$ being the center of the score, corresponding to using $0$ as the mean for the prior for $\mu$.

The ordered logistic model can be written as:

$\begin{aligned} & y_i= \begin{cases}1 & x_i+\epsilon_i \leq c_1 \\ 2 & c_1<x_i+\epsilon_i \leq c_2 \\ \ldots & \\ 7 & x_i+\epsilon_i>c_6\end{cases} \\ & \epsilon_i \sim \operatorname{Logistic}(0,1)\end{aligned}$

## 1.2 Model Implementation

```{r}
# keep ONLY rows where happiness is observed
dat_clean <- dat %>%
  filter(!is.na(happy)) %>%
  mutate(happy = as.integer(happy))

table(dat_clean$happy)
sum(is.na(dat_clean$happy))

y <- dat_clean$happy

stan_data <- list(
  N = length(y),
  y = y
)
```

### 1.2.1 HMC with Stan

```{r}
# -----------------------------
# 4) Compile Stan model
dir.create("stan_out", showWarnings = FALSE)
# -----------------------------
mod <- cmdstan_model("naive_model.stan")

# -----------------------------
# 5) Fit with HMC / NUTS
# -----------------------------
fit_hmc <- mod$sample(
  data = stan_data,
  iter_warmup = 1000,
  iter_sampling = 2000,
  chains = 1,
  refresh = 500,
  output_dir = "stan_out"
  
)

# -----------------------------
# 6) Summaries
# -----------------------------
fit_hmc$summary(c("mu", "sigma"))
```

### 1.2.2 VI with Stan

```{r}
# -----------------------------
mod <- cmdstan_model("naive_model.stan")

# -----------------------------
# 5) Fit with VI
# -----------------------------
fit_vi = mod$variational(
  output_samples = 3000,
  refresh = 500,
  output_dir = "stan_out",
  algorithm = "meanfield",
  data = stan_data
)
```

```{r}
fit_vi$summary(c("mu", "sigma"))
```

### 1.2.3 Comparison between VI and HMC

we quantify the discrepancy between HMC and VI using standardized difference in posterior means and ratio of posterior standard deviation.

```{r}
hmc <- fit_hmc$summary(c("mu","sigma"))
vi  <- fit_vi$summary(c("mu","sigma"))
diff_mu <- abs(hmc$mean[1] - vi$mean[1]) / hmc$sd[1]
ratio_mu <- vi$sd[1] / hmc$sd[1]

diff_sigma <- abs(hmc$mean[2] - vi$mean[2]) / hmc$sd[2]
ratio_sigma <- vi$sd[2] / hmc$sd[2]
diff_mu
ratio_mu
diff_sigma
ratio_sigma
```

```{r}
library(loo)

# HMC
log_lik_hmc <- fit_hmc$draws("log_lik", format = "matrix")
loo_hmc <- loo(log_lik_hmc)

# VI
log_lik_vi <- fit_vi$draws("log_lik", format = "matrix")
loo_vi <- loo(log_lik_vi)

loo_hmc
loo_vi
```

```{r}
loo_compare(loo_hmc, loo_vi)
```

from the results the prediction accuracy differ by a lot.

## 1.3 VI Model Diagnostic

```{r}
# -----------------------------
# 1) Hold out the last observation
# -----------------------------
y_full <- stan_data$y
y_test <- y_full[length(y_full)]
y_train <- y_full[-length(y_full)]

stan_data_loo_last <- list(
  N = length(y_train),
  y = y_train
)

# -----------------------------
# 2) Compile model
# -----------------------------
mod <- cmdstan_model("naive_model.stan")

# -----------------------------
# 3) Fit with VI on n-1 data
# -----------------------------
fit_vi <- mod$variational(
  output_samples = 2000,
  refresh = 500,
  output_dir = "stan_out",
  algorithm = "meanfield",
  data = stan_data_loo_last
)

# -----------------------------
# 4) Extract posterior predictive draws for y_new
# -----------------------------
y_new_draws <- as.numeric(fit_vi$draws("y_new"))

# -----------------------------
# 5) Compute a 99% predictive credible interval
# -----------------------------
ci <- quantile(y_new_draws, probs = c(0.025, 0.975))

lower <- max(1, ceiling(ci[1]))
upper <- min(7, floor(ci[2]))

c(lower = lower, upper = upper)

# -----------------------------
# 6) Check whether held-out y_n is inside the interval
# -----------------------------
y_test
inside_interval <- (y_test >= lower) && (y_test <= upper)
inside_interval
```

There is no warning spotted.

## 2. Intermediate Model (with time structure)

## 2.1 Model setup

For each observation $i$ :

-   $y_i \in\{1,2,3,4,5,6,7\}$ : observed happiness rating (ordinal)

-   $x_i \in \mathbb{R}$ : latent continuous happiness

**Latent Dynamic:**

Instead of assuming independence, we model $x_i$ conditionally on its past values:

$$
x_i \mid x_{i-1}, x_{i-5} \sim \mathcal{N}\left(\mu_i, \sigma^2\right)
$$

where the conditional mean is given by a regression structure:

$$
\mu_i=\alpha+\beta_1 x_{i-1}+\beta_5 x_{i-5}
$$

This introduces:

-   short-term dependence via $x_{i-1}$

-   longer-term dependence via $x_{i-5}$

**Initial Condition\
**

For the first few observations where lags are not available (i.e., $i \leq 5$ ), we use:

$$
x_i \sim \mathcal{N}\left(\mu_0, \sigma^2\right)
$$

with $\mu_0=\alpha$, ensuring a consistent baseline.

We place priors on the regression parameters to control the influence of temporal dependence:

$$
\begin{aligned}
& \alpha \sim \mathcal{N}\left(0,1.5^2\right) \\
& \beta_1 \sim \mathcal{N}(0,1), \quad \beta_5 \sim \mathcal{N}(0,1) \\
& \sigma \sim \text { Exponential }(1)
\end{aligned}
$$

Interpretation: - $\alpha$ : baseline latent happiness level - $\beta_{1 \text { : }}$ immediate persistence (short-term effect) - $\beta_5$ : delayed persistence (long-term effect) - $\sigma$ : variability not explained by temporal structure

**Likelihood**

Same as before

## 2.2 Model Implementation

Since simply removing missing observation would distort the temporal structure of the data by altering the spacing between time points. Therefore we retain all time indices and

```{r}
dat <- read_tsv("Data/Backup/backup.tsv", show_col_types = FALSE) %>%
  arrange(id, day, beep) %>%
  mutate(
    happy = as.integer(happy),
    row_index = row_number()
  )

# -----------------------------
# 1) Build Stan data for the intermediate model
# -----------------------------
N <- nrow(dat)

is_obs <- !is.na(dat$happy)
obs_index <- which(is_obs)
y_obs <- dat$happy[is_obs]
N_obs <- length(y_obs)

has_lag1 <- rep(0L, N)
lag1_index <- rep(0L, N)

has_lag5 <- rep(0L, N)
lag5_index <- rep(0L, N)

for (i in seq_len(N)) {
  # lag 1: previous row must belong to same participant
  if (i > 1 && dat$id[i] == dat$id[i - 1]) {
    has_lag1[i] <- 1L
    lag1_index[i] <- i - 1L
  }

  # lag 5: 5 rows before must belong to same participant
  if (i > 5 && dat$id[i] == dat$id[i - 5]) {
    has_lag5[i] <- 1L
    lag5_index[i] <- i - 5L
  }
}

stan_data_intermediate <- list(
  N = N,
  N_obs = N_obs,
  y_obs = y_obs,
  obs_index = obs_index,
  has_lag1 = has_lag1,
  lag1_index = lag1_index,
  has_lag5 = has_lag5,
  lag5_index = lag5_index)
```

## 2.3 VI Model rerun

```{r}
# -----------------------------
# 2) Compile model
# -----------------------------
mod_intermediate <- cmdstan_model("intermediate_model.stan")

# -----------------------------
# 3) Fit with VI
# -----------------------------
dir.create("stan_out", showWarnings = FALSE)

fit_vi_intermediate <- mod_intermediate$variational(
  data = stan_data_intermediate,
  algorithm = "meanfield",
  output_samples = 3000,
  refresh = 500,
  output_dir = "stan_out"
)
```

```{r}
# -----------------------------
# 1) Hold out the last observation
# -----------------------------
y_full <- stan_data_intermediate$y
y_test <- y_full[length(y_full)]
y_train <- y_full[-length(y_full)]

stan_data_loo_last <- list(
  N = length(y_train),
  y = y_train
)

# -----------------------------
# 2) Compile model
# -----------------------------
mod <- cmdstan_model("naive_model.stan")

# -----------------------------
# 3) Fit with VI on n-1 data
# -----------------------------
fit_vi <- mod$variational(
  output_samples = 2000,
  refresh = 500,
  output_dir = "stan_out",
  algorithm = "meanfield",
  data = stan_data_loo_last
)

# -----------------------------
# 4) Extract posterior predictive draws for y_new
# -----------------------------
y_new_draws <- as.numeric(fit_vi$draws("y_new"))

# -----------------------------
# 5) Compute a 99% predictive credible interval
# -----------------------------
ci <- quantile(y_new_draws, probs = c(0.025, 0.975))

lower <- max(1, ceiling(ci[1]))
upper <- min(7, floor(ci[2]))

c(lower = lower, upper = upper)

# -----------------------------
# 6) Check whether held-out y_n is inside the interval
# -----------------------------
y_test
inside_interval <- (y_test >= lower) && (y_test <= upper)
inside_interval
```

## 2.4 Vi & HMC Comparison

```{r}
# -----------------------------
# 4) Compile Stan model
dir.create("stan_out", showWarnings = FALSE)
# -----------------------------
mod <- cmdstan_model("intermediate_model.stan")

# -----------------------------
# 5) Fit with HMC / NUTS
# -----------------------------
fit_hmc <- mod$sample(
  data = stan_data_intermediate,
  iter_warmup = 10,
  iter_sampling = 20,
  chains = 1,
  refresh = 5,
  output_dir = "stan_out",
  max_treedepth = 15 
)
```

```{r}
library(loo)

# HMC
log_lik_hmc <- fit_hmc$draws("log_lik", format = "matrix")
loo_hmc <- loo(log_lik_hmc)

# VI
log_lik_vi <- fit_vi_intermediate$draws("log_lik", format = "matrix")
loo_vi <- loo(log_lik_vi)

loo_hmc
loo_vi
```

```{r}
loo_compare(loo_hmc, loo_vi)
```

## 2.5 Model fitting invariance

```{r}
# 4) Compile Stan model
dir.create("stan_out", showWarnings = FALSE)
# -----------------------------
mod <- cmdstan_model("intermediate_model.stan")

# -----------------------------
# 5) Fit with HMC / NUTS
# -----------------------------
fit_hmc <- mod$sample(
  data = stan_data_intermediate,
  seed = 405,
  iter_warmup = 1000,
  iter_sampling = 2000,
  chains = 2,
  refresh = 500,
  output_dir = "stan_out",
  max_treedepth = 15
)
```

```{r}
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(bayesplot))

mcmc_trace(fit_hmc$draws("alpha")) + theme_minimal()
mcmc_rank_hist(fit_hmc$draws("alpha"))
```

## 3 Advanced Model

Latent Dynamic with Covariates

We extend the intermediate model by incorporating both:

-   temporal dependence

-   lagged emotional covariates

-   $x_i \mid x_{i-1}, x_{i-5}, \mathbf{Z}_{i-1} \sim \mathcal{N}\left(\mu_i, \sigma^2\right)$ where: $\mu_i=\alpha+\beta_1 x_{i-1}+\beta_5 x_{i-5}+\gamma_1$ enthusiastic $_{i-1}+\gamma_2$ pleased $_{i-1}+\gamma_3$ relaxed $_{i-1}$

Initial Condition

For early observations where lags are unavailable (i.e., $i \leq 5$ ):

$$
x_i \sim \mathcal{N}\left(\mu_0, \sigma^2\right)
$$

where:

$$
\mu_0=\alpha
$$

ensuring a consistent baseline.

Prior Specification

We place weakly informative priors:

$$
\begin{aligned}
& \alpha \sim \mathcal{N}\left(0,1.5^2\right) \\
& \beta_1 \sim \mathcal{N}(0,1), \quad \beta_5 \sim \mathcal{N}(0,1) \\
& \gamma_1 \sim \mathcal{N}(0,1), \quad \gamma_2 \sim \mathcal{N}(0,1), \quad \gamma_3 \sim \mathcal{N}(0,1) \\
& \sigma \sim \text { Exponential }(1)
\end{aligned}
$$

## 3.2 Model Implementation

```{r}

# -----------------------------
dat <- read_tsv("Data/Backup/backup.tsv", show_col_types = FALSE) %>%
  arrange(id, day, beep) %>%
  mutate(
    happy = as.integer(happy),
    row_index = row_number()
  )

# -----------------------------
# 1) Standardize exogenous predictors using observed values only
# -----------------------------
scale_to_numeric <- function(x) {
  m <- mean(x, na.rm = TRUE)
  s <- sd(x, na.rm = TRUE)
  (x - m) / s
}

dat <- dat %>%
  mutate(
    enthusiastic_std = scale_to_numeric(enthusiastic),
    pleased_std      = scale_to_numeric(pleased),
    relaxed_std      = scale_to_numeric(relaxed)
  )

# -----------------------------
# 2) Basic observed-y objects
# -----------------------------
N <- nrow(dat)

is_obs <- !is.na(dat$happy)
obs_index <- which(is_obs)
y_obs <- dat$happy[is_obs]
N_obs <- length(y_obs)

# -----------------------------
# 3) Build latent lag structure
# -----------------------------
has_lag1 <- rep(0L, N)
lag1_index <- rep(0L, N)

has_lag5 <- rep(0L, N)
lag5_index <- rep(0L, N)

for (i in seq_len(N)) {
  # lag 1: previous row must be same participant
  if (i > 1 && dat$id[i] == dat$id[i - 1]) {
    has_lag1[i] <- 1L
    lag1_index[i] <- i - 1L
  }

  # lag 5: five rows before must be same participant
  if (i > 5 && dat$id[i] == dat$id[i - 5]) {
    has_lag5[i] <- 1L
    lag5_index[i] <- i - 5L
  }
}

# -----------------------------
# 4) Build lagged exogenous predictors
#    Use previous-row values only if same participant and observed
# -----------------------------
enth_lag1 <- rep(0, N)
pleased_lag1 <- rep(0, N)
relaxed_lag1 <- rep(0, N)

has_enth_lag1 <- rep(0L, N)
has_pleased_lag1 <- rep(0L, N)
has_relaxed_lag1 <- rep(0L, N)

for (i in seq_len(N)) {
  if (i > 1 && dat$id[i] == dat$id[i - 1]) {
    if (!is.na(dat$enthusiastic_std[i - 1])) {
      enth_lag1[i] <- dat$enthusiastic_std[i - 1]
      has_enth_lag1[i] <- 1L
    }
    if (!is.na(dat$pleased_std[i - 1])) {
      pleased_lag1[i] <- dat$pleased_std[i - 1]
      has_pleased_lag1[i] <- 1L
    }
    if (!is.na(dat$relaxed_std[i - 1])) {
      relaxed_lag1[i] <- dat$relaxed_std[i - 1]
      has_relaxed_lag1[i] <- 1L
    }
  }
}

# -----------------------------
# 5) Final Stan data list
# -----------------------------
stan_data_model3 <- list(
  N = N,
  N_obs = N_obs,
  y_obs = y_obs,
  obs_index = obs_index,
  has_lag1 = has_lag1,
  lag1_index = lag1_index,
  has_lag5 = has_lag5,
  lag5_index = lag5_index,
  enth_lag1 = enth_lag1,
  pleased_lag1 = pleased_lag1,
  relaxed_lag1 = relaxed_lag1,
  has_enth_lag1 = has_enth_lag1,
  has_pleased_lag1 = has_pleased_lag1,
  has_relaxed_lag1 = has_relaxed_lag1
)

# -----------------------------
# 6) Sanity checks
# -----------------------------
stopifnot(length(y_obs) == N_obs)
stopifnot(length(obs_index) == N_obs)
stopifnot(length(has_lag1) == N)
stopifnot(length(lag1_index) == N)
stopifnot(length(has_lag5) == N)
stopifnot(length(lag5_index) == N)
stopifnot(length(enth_lag1) == N)
stopifnot(length(pleased_lag1) == N)
stopifnot(length(relaxed_lag1) == N)
```

```{r}
dir.create("stan_out", showWarnings = FALSE)

mod_model3 <- cmdstan_model("advanced_model.stan")

fit_hmc_model3 <- mod_model3$sample(
  data = stan_data_model3,
  chains = 1,
  iter_warmup = 100,
  iter_sampling = 200,
  output_dir = "stan_out"
)
```

```{r}
fit_vi_model3 <- mod_model3$variational(
  data = stan_data_model3,
  algorithm = "meanfield",
  output_samples = 3000,
  refresh = 500,
  output_dir = "stan_out"
)
```

```{r}
library(loo)

# HMC
log_lik_hmc <- fit_hmc_model3$draws("log_lik", format = "matrix")
loo_hmc <- loo(log_lik_hmc)

# VI
log_lik_vi <- fit_vi_model3$draws("log_lik", format = "matrix")
loo_vi <- loo(log_lik_vi)

loo_hmc
loo_vi
```

```{r}
loo_compare(loo_hmc, loo_vi)
```

```{r}
dir.create("stan_out", showWarnings = FALSE)

mod_model3 <- cmdstan_model("advanced_model.stan")

fit_hmc_model3 <- mod_model3$sample(
  data = stan_data_model3,
  chains = 2,
  iter_warmup = 1000,
  iter_sampling = 2000,
  refresh = 25,
  output_dir = "stan_out"
)
```

```{r}
mcmc_trace(fit_hmc_model3$draws("alpha_raw")) + theme_minimal()
mcmc_rank_hist(fit_hmc_model3$draws("alpha_raw
```
