---
title: "Variational Inference Model fitting"
author: "Marc Shi"
format: pdf
editor: visual
---

```{r}
library(cmdstanr)
library(readr)
library(dplyr)
set.seed(405)

```

# Data Importing & Cleaning

```{r}

dat <- read_tsv("Data/Backup/backup.tsv", show_col_types = FALSE)
```

from the dataset, we see that there are 7350 observations, which comes from 105 participants × 14 days $\times 5$ beeps per day $=7350$. Therefore for each participants, we have $14 \times 5=70$ timepoints per participant.

# Model Fitting

## 1. Naive Model (Baseline)

For the naive model, we use the **independent latent variable model with ordered logistic likelihood.**

## 1.1 Model set up

For each observation $i$ :

-   $y_i \in\{1,2,3,4,5,6,7\}$ : observed happiness rating (ordinal)

-   $x_i \in \mathbb{R}$ : latent continuous happiness

**Prior:** we assume $x_i \sim \operatorname{iid} \mathcal{N}\left(\mu, \sigma^2\right)$, where each latent happiness level is independent across time (within-participant) and between participants and is drawn from a same distribution with parameter. Since there is only one pbserved score $y_i$ connected to $x_i$ we rom the prior, we will be hierarchical and apply the hyper parameters as below:

$\mu \sim \mathcal{N}\left(0,1.5^2\right)$ and $\sigma \sim$ Exponential $(1)$

**Likelihood (Ordered Logistic Model)**

$$
y_i \mid x_i \sim \operatorname{Ordered} \operatorname{Logistic}\left(x_i, c\right)
$$

where: . $c=\left(c_1, c_2, c_3, c_4, c_5, c_6\right)$ are cutpoints with

$c = (-2.5,-1.5,-0.5,0.5,1.5,2.5)$, thus make $0$ being the center of the score, corresponding to using $0$ as the mean for the prior for $\mu$.

The ordered logistic model can be written as:

$\begin{aligned} & y_i= \begin{cases}1 & x_i+\epsilon_i \leq c_1 \\ 2 & c_1<x_i+\epsilon_i \leq c_2 \\ \ldots & \\ 7 & x_i+\epsilon_i>c_6\end{cases} \\ & \epsilon_i \sim \operatorname{Logistic}(0,1)\end{aligned}$

## 1.2 Model Implementation

```{r}
# keep ONLY rows where happiness is observed
dat_clean <- dat %>%
  filter(!is.na(happy)) %>%
  mutate(happy = as.integer(happy))

table(dat_clean$happy)
sum(is.na(dat_clean$happy))

y <- dat_clean$happy

stan_data <- list(
  N = length(y),
  y = y
)
```

### 1.2.1 HMC with Stan

```{r}

dir.create("stan_out", showWarnings = FALSE)
# -----------------------------
mod <- cmdstan_model("naive_model.stan")


fit_hmc <- mod$sample(
  data = stan_data,
  chains = 2,
  iter_warmup = 1000,
  iter_sampling = 2000,
  refresh = 500,
  output_dir = "stan_out"
)

# -----------------------------
# 6) Summaries
# -----------------------------
fit_hmc$summary(c("mu", "sigma"))
```

```{r}
fit_hmc$summary(c("mu", "sigma","y_new"))
fit_hmc$diagnostic_summary()
```

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(bayesplot)
  library(gridExtra)
  library(grid)
})

p1<-mcmc_trace(fit_hmc$draws("mu")) + theme_minimal()

```

```{r}
p2<-mcmc_rank_hist(fit_hmc$draws("mu")) 
desktop <- file.path(Sys.getenv("USERPROFILE"), "Desktop")   # Windows

pdf(file.path(desktop, "mu_diagnostics.pdf"), width = 12, height = 5)
grid.arrange(p1, p2, ncol = 2)
dev.off()

```

### 1.2.2 VI with Stan

```{r}
# -----------------------------
mod <- cmdstan_model("naive_model.stan")

# -----------------------------
# 5) Fit with VI
# -----------------------------
fit_vi = mod$variational(
  output_samples = 2000,
  refresh = 500,
  output_dir = "stan_out",
  algorithm = "meanfield",
  data = stan_data
)
```

```{r}
fit_vi$summary(c("mu", "sigma","y_new"))
```

### 1.2.3 Comparison between VI and HMC

we quantify the discrepancy between HMC and VI using standardized difference in posterior means and ratio of posterior standard deviation.

```{r}
hmc <- fit_hmc$summary(c("mu","sigma"))
vi  <- fit_vi$summary(c("mu","sigma"))
diff_mu <- abs(hmc$mean[1] - vi$mean[1]) / hmc$sd[1]
ratio_mu <- vi$sd[1] / hmc$sd[1]

diff_sigma <- abs(hmc$mean[2] - vi$mean[2]) / hmc$sd[2]
ratio_sigma <- vi$sd[2] / hmc$sd[2]
diff_mu
ratio_mu
diff_sigma
ratio_sigma
```

```{r}
library(loo)

# HMC
log_lik_hmc <- fit_hmc$draws("log_lik", format = "matrix")
loo_hmc <- loo(log_lik_hmc)

# VI
log_lik_vi <- fit_vi$draws("log_lik", format = "matrix")
loo_vi <- loo(log_lik_vi)

loo_hmc
loo_vi
```

```{r}
loo_compare(loo_hmc, loo_vi)
```

from the results the prediction accuracy differ by a lot.

## 1.3 VI Model Diagnostic

```{r}

# 1)Hold out the last observation

y_full <- stan_data$y
y_test <- y_full[length(y_full)]
y_train <- y_full[-length(y_full)]

stan_data_loo_last <- list(
  N = length(y_train),
  y = y_train
)


# 2) Compile model
# -----------------------------
mod <- cmdstan_model("naive_model.stan")


# 3) Fit

fit_vi <- mod$variational(
  output_samples = 2000,
  refresh = 500,
  output_dir = "stan_out",
  algorithm = "meanfield",
  data = stan_data_loo_last
)

# 4)Extract posterior predictive draws for y_new

y_new_draws <- as.numeric(fit_vi$draws("y_new"))


# 5) Compute a 99% predictive credible interval

ci <- quantile(y_new_draws, probs = c(0.005, 0.995))

lower <- max(1, ceiling(ci[1]))
upper <- min(7, floor(ci[2]))

c(lower = lower, upper = upper)


# 6) Check whether held-out y_n is inside the interval

y_test
inside_interval <- (y_test >= lower) && (y_test <= upper)
inside_interval
```

There is no warning spotted.

## 1.4 HMC Model Diagnostic

```{r}


# 1) Hold out the last observation

y_full <- stan_data$y
y_test <- y_full[length(y_full)]
y_train <- y_full[-length(y_full)]

stan_data_loo_last <- list(
  N = length(y_train),
  y = y_train
)


# 2) Compile model

mod <- cmdstan_model("naive_model.stan")


# 3) Fit with HMC on n-1 dat

fit_hmc_loo <- mod$sample(
  data = stan_data_loo_last,
  iter_warmup = 1000,
  iter_sampling = 2000,
  chains = 1,
  refresh = 500,
  seed = 405,
  output_dir = "stan_out"
)


# 4) Extract posterior predictive draws for y_new

y_new_draws <- fit_hmc_loo$draws("y_new", format = "matrix")[, 1]


# 5) Compute a 95% predictive credible interval

ci <- quantile(y_new_draws, probs = c(0.025, 0.975))

lower <- max(1, ceiling(ci[1]))
upper <- min(7, floor(ci[2]))

c(lower = lower, upper = upper)


# 6) Check whether held-out y_n is inside the interval

y_test
inside_interval <- (y_test >= lower) && (y_test <= upper)
inside_interval
```

## 1.5 Exact invariance Test

```{r}


cuts_fixed <- c(-2.5, -1.5, -0.5, 0.5, 1.5, 2.5)

# ---------- ordered logistic helpers ----------
ordered_logistic_probs <- function(eta, cuts = cuts_fixed) {
  K <- length(cuts) + 1
  p <- numeric(K)

  p[1] <- plogis(cuts[1] - eta)
  if (K > 2) {
    for (k in 2:(K - 1)) {
      p[k] <- plogis(cuts[k] - eta) - plogis(cuts[k - 1] - eta)
    }
  }
  p[K] <- 1 - plogis(cuts[K - 1] - eta)

  pmax(p, 1e-12)
}

rordered_logistic1 <- function(eta, cuts = cuts_fixed) {
  sample.int(length(cuts) + 1, size = 1, prob = ordered_logistic_probs(eta, cuts))
}

log_ordered_logistic_vec <- function(y, eta, cuts = cuts_fixed) {
  probs <- vapply(
    seq_along(y),
    function(i) ordered_logistic_probs(eta[i], cuts)[y[i]],
    numeric(1)
  )
  sum(log(pmax(probs, 1e-12)))
}

# ---------- forward simulator from the JOINT model ----------
forward_naive <- function(N = 30, cuts = cuts_fixed) {
  mu <- rnorm(1, mean = 0, sd = 1.5)
  sigma <- rexp(1, rate = 1)
  z <- rnorm(N)
  x <- mu + sigma * z
  y <- vapply(x, rordered_logistic1, integer(1), cuts = cuts)

  list(
    mu = mu,
    ell = log(sigma),
    z = z,
    y = y
  )
}

# ---------- log joint density in (mu, ell, z, y) parameterization ----------
log_joint_naive <- function(mu, ell, z, y, cuts = cuts_fixed) {
  sigma <- exp(ell)
  x <- mu + sigma * z

  dnorm(mu, mean = 0, sd = 1.5, log = TRUE) +
    dexp(sigma, rate = 1, log = TRUE) +
    ell +
    sum(dnorm(z, mean = 0, sd = 1, log = TRUE)) +
    log_ordered_logistic_vec(y, x, cuts)
}

# ---------- one fixed Metropolis-within-Gibbs kernel ----------
stationary_mcmc_naive <- function(
    n_iter,
    N = 30,
    cuts = cuts_fixed,
    prop_sd_mu = 0.30,
    prop_sd_ell = 0.20,
    prop_sd_z = 0.40
) {
  state <- forward_naive(N = N, cuts = cuts)
  y <- state$y

  if (n_iter == 0) return(state)

  current_lp <- log_joint_naive(state$mu, state$ell, state$z, y, cuts)

  for (iter in seq_len(n_iter)) {

    mu_prop <- rnorm(1, mean = state$mu, sd = prop_sd_mu)
    lp_prop <- log_joint_naive(mu_prop, state$ell, state$z, y, cuts)
    if (log(runif(1)) < (lp_prop - current_lp)) {
      state$mu <- mu_prop
      current_lp <- lp_prop
    }

    ell_prop <- rnorm(1, mean = state$ell, sd = prop_sd_ell)
    lp_prop <- log_joint_naive(state$mu, ell_prop, state$z, y, cuts)
    if (log(runif(1)) < (lp_prop - current_lp)) {
      state$ell <- ell_prop
      current_lp <- lp_prop
    }

    for (n in seq_along(state$z)) {
      z_prop <- state$z
      z_prop[n] <- rnorm(1, mean = state$z[n], sd = prop_sd_z)

      lp_prop <- log_joint_naive(state$mu, state$ell, z_prop, y, cuts)
      if (log(runif(1)) < (lp_prop - current_lp)) {
        state$z <- z_prop
        current_lp <- lp_prop
      }
    }
  }

  state
}

# ---------- choose scalar summaries to compare ----------
extract_summary <- function(state, summary = c("mu", "sigma", "xbar", "x1")) {
  summary <- match.arg(summary)

  sigma <- exp(state$ell)
  x <- state$mu + sigma * state$z

  switch(
    summary,
    mu = state$mu,
    sigma = sigma,
  )
}

# ---------- exact invariance test ----------
exact_invariance_test_naive <- function(
    n_rep = 300,
    n_iter = 50,
    N = 30,
    summaries = c("mu", "sigma", "xbar", "x1")
) {
  out <- lapply(summaries, function(s) {
    forward_only <- replicate(
      n_rep,
      extract_summary(stationary_mcmc_naive(n_iter = 0, N = N), s)
    )

    with_mcmc <- replicate(
      n_rep,
      extract_summary(stationary_mcmc_naive(n_iter = n_iter, N = N), s)
    )

    ks <- ks.test(forward_only, with_mcmc)

    data.frame(
      summary = s,
      D_statistic = unname(ks$statistic),
      p_value = ks$p.value,
      row.names = NULL
    )
  })

  do.call(rbind, out)
}

# testtttttttttttttttttttttttttttttttttttttt
eit_results <- exact_invariance_test_naive(
  n_rep = 200, #300
  n_iter = 30, #50
  N = 20, #30
  summaries = c("mu", "sigma")
)

eit_results


```

## 1.6 ESS

```{r}

ess_table <- fit_hmc$summary(
  NULL,
  c("mean", "sd", "mcse_mean", "ess_bulk", "ess_tail", "rhat")
)

ess_table[ess_table$variable %in% c("mu", "sigma"),
          c("variable", "mean", "sd", "mcse_mean", "ess_bulk", "ess_tail", "rhat")]
```

## 2. Intermediate Model (with time structure)

## 2.1 Model setup

For each observation $i$ :

-   $y_i \in\{1,2,3,4,5,6,7\}$ : observed happiness rating (ordinal)

-   $x_i \in \mathbb{R}$ : latent continuous happiness

**Latent Dynamic:**

Instead of assuming independence, we model $x_i$ conditionally on its past values:

$$
x_i \mid x_{i-1}, x_{i-5} \sim \mathcal{N}\left(\mu_i, \sigma^2\right)
$$

where the conditional mean is given by a regression structure:

$$
\mu_i=\alpha+\beta_1 x_{i-1}+\beta_5 x_{i-5}
$$

This introduces:

-   short-term dependence via $x_{i-1}$

-   longer-term dependence via $x_{i-5}$

**Initial Condition\
**

For the first few observations where lags are not available (i.e., $i \leq 5$ ), we use:

$$
x_i \sim \mathcal{N}\left(\mu_0, \sigma^2\right)
$$

with $\mu_0=\alpha$, ensuring a consistent baseline.

We place priors on the regression parameters to control the influence of temporal dependence:

$$
\begin{aligned}
& \alpha \sim \mathcal{N}\left(0,1.5^2\right) \\
& \beta_1 \sim \mathcal{N}(0,1), \quad \beta_5 \sim \mathcal{N}(0,1) \\
& \sigma \sim \text { Exponential }(1)
\end{aligned}
$$

Interpretation: - $\alpha$ : baseline latent happiness level - $\beta_{1 \text { : }}$ immediate persistence (short-term effect) - $\beta_5$ : delayed persistence (long-term effect) - $\sigma$ : variability not explained by temporal structure

**Likelihood**

Same as before

## 2.2 Model Implementation

Since simply removing missing observation would distort the temporal structure of the data by altering the spacing between time points. Therefore we retain all time indices and

```{r}
dat <- read_tsv("Data/Backup/backup.tsv", show_col_types = FALSE) %>%
  arrange(id, day, beep) %>%
  mutate(
    happy = as.integer(happy),
    row_index = row_number()
  )


# 1) Build Stan data for the intermediate model

N <- nrow(dat)

is_obs <- !is.na(dat$happy)
obs_index <- which(is_obs)
y_obs <- dat$happy[is_obs]
N_obs <- length(y_obs)

has_lag1 <- rep(0L, N)
lag1_index <- rep(0L, N)

has_lag5 <- rep(0L, N)
lag5_index <- rep(0L, N)

for (i in seq_len(N)) {
  # lag 1: previous row must belong to same participant
  if (i > 1 && dat$id[i] == dat$id[i - 1]) {
    has_lag1[i] <- 1L
    lag1_index[i] <- i - 1L
  }

  # lag 5: 5 rows before must belong to same participant
  if (i > 5 && dat$id[i] == dat$id[i - 5]) {
    has_lag5[i] <- 1L
    lag5_index[i] <- i - 5L
  }
}

stan_data_intermediate <- list(
  N = N,
  N_obs = N_obs,
  y_obs = y_obs,
  obs_index = obs_index,
  has_lag1 = has_lag1,
  lag1_index = lag1_index,
  has_lag5 = has_lag5,
  lag5_index = lag5_index)
```

```{r}

# 4) Compile
dir.create("stan_out", showWarnings = FALSE)
# -----------------------------
mod <- cmdstan_model("intermediate_model.stan")


# 5) Fit with HMC / NUTS

fit_hmc <- mod$sample(
  data = stan_data_intermediate,
  seed = 405,
  iter_warmup = 500,
  iter_sampling = 1000,
  chains = 2,
  refresh = 250,
  output_dir = "stan_out",
  max_treedepth = 15
)
```

## 1.3 HMC Model Diagnostic

```{r}


# 1) Hold out the last observation

y_full <- stan_data_intermediate$y
y_test <- y_full[length(y_full)]
y_train <- y_full[-length(y_full)]

stan_data_loo_last <- list(
  N = length(y_train),
  y = y_train
)


# 2) Compile model

mod <- cmdstan_model("intermediate_model.stan")


# 3) Fit with HMC on n-1 dat

fit_hmc_loo <- mod$sample(
  data = stan_data_loo_last,
  iter_warmup = 1000,
  iter_sampling = 2000,
  chains = 1,
  refresh = 500,
  seed = 405,
  output_dir = "stan_out"
)


# 4) Extract posterior predictive draws for y_new

y_new_draws <- fit_hmc_loo$draws("y_new", format = "matrix")[, 1]


# 5) Compute a 95% predictive credible interval

ci <- quantile(y_new_draws, probs = c(0.025, 0.975))

lower <- max(1, ceiling(ci[1]))
upper <- min(7, floor(ci[2]))

c(lower = lower, upper = upper)


# 6) Check whether held-out y_n is inside the interval

y_test
inside_interval <- (y_test >= lower) && (y_test <= upper)
inside_interval

```

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(bayesplot)
  library(gridExtra)
  library(grid)
})

p1<-mcmc_trace(fit_hmc$draws("sigma")) + theme_minimal()

```

```{r}
p2<-mcmc_rank_hist(fit_hmc$draws("sigma")) 
desktop <- file.path(Sys.getenv("USERPROFILE"), "Desktop")   # Windows

pdf(file.path(desktop, "intermediate_sigma_diagnostics.pdf"), width = 12, height = 5)
grid.arrange(p1, p2, ncol = 2)
dev.off()

```

\

\

## 1.3 Exact Invariance Test

```{r}



cuts_fixed <- c(-2.5, -1.5, -0.5, 0.5, 1.5, 2.5)

# ordered-logistic helpers
ordered_logistic_probs <- function(eta, cuts = cuts_fixed) {
  p <- numeric(length(cuts) + 1)
  p[1] <- plogis(cuts[1] - eta)
  for (k in 2:length(cuts)) {
    p[k] <- plogis(cuts[k] - eta) - plogis(cuts[k - 1] - eta)
  }
  p[length(cuts) + 1] <- 1 - plogis(cuts[length(cuts)] - eta)
  pmax(p, 1e-12)
}

rordered_logistic1 <- function(eta, cuts = cuts_fixed) {
  sample.int(length(cuts) + 1, 1, prob = ordered_logistic_probs(eta, cuts))
}

log_ordered_logistic_vec <- function(y, eta, cuts = cuts_fixed) {
  probs <- mapply(function(yy, ee) ordered_logistic_probs(ee, cuts)[yy], y, eta)
  sum(log(pmax(probs, 1e-12)))
}

# recursive latent trajectory for the intermediate model
make_x <- function(alpha, beta1, beta5, ell, z) {
  sigma <- exp(ell)
  N <- length(z)
  x <- numeric(N)

  for (i in seq_len(N)) {
    mu_i <- alpha
    if (i > 1) mu_i <- mu_i + beta1 * x[i - 1]
    if (i > 5) mu_i <- mu_i + beta5 * x[i - 5]
    x[i] <- mu_i + sigma * z[i]
    if (!is.finite(x[i])) return(NULL)
  }

  x
}

# forward simulator
forward_intermediate <- function(N = 20, cuts = cuts_fixed) {
  repeat {
    alpha <- rnorm(1, 0, 1.5)
    beta1 <- rnorm(1, 0, 1)
    beta5 <- rnorm(1, 0, 1)
    sigma <- rexp(1, 1)
    z <- rnorm(N)

    x <- make_x(alpha, beta1, beta5, log(sigma), z)
    if (!is.null(x)) {
      y <- sapply(x, rordered_logistic1, cuts = cuts)
      return(list(alpha = alpha, beta1 = beta1, beta5 = beta5,
                  ell = log(sigma), z = z, y = y))
    }
  }
}

# joint density
joint_intermediate <- function(state, y, cuts = cuts_fixed) {
  x <- make_x(state$alpha, state$beta1, state$beta5, state$ell, state$z)
  if (is.null(x)) return(-Inf)

  sigma <- exp(state$ell)

  dnorm(state$alpha, 0, 1.5, log = TRUE) +
    dnorm(state$beta1, 0, 1, log = TRUE) +
    dnorm(state$beta5, 0, 1, log = TRUE) +
    dexp(sigma, 1, log = TRUE) +
    state$ell +
    sum(dnorm(state$z, 0, 1, log = TRUE)) +
    log_ordered_logistic_vec(y, x, cuts)
}

# summary extractor
extract_summary_intermediate <- function(state, summary = "alpha") {
  x <- make_x(state$alpha, state$beta1, state$beta5, state$ell, state$z)
  sigma <- exp(state$ell)

  switch(
    summary,
    alpha = state$alpha,
    beta1 = state$beta1,
    beta5 = state$beta5,
    sigma = sigma,
    xbar = mean(x),
    x6 = x[6]
  )
}

# one MH update helper
mh_update <- function(state, y, logp_current, proposal_fun) {
  proposal <- proposal_fun(state)
  logp_prop <- joint_intermediate(proposal, y)

  if (log(runif(1)) < (logp_prop - logp_current)) {
    list(state = proposal, logp = logp_prop)
  } else {
    list(state = state, logp = logp_current)
  }
}

# forward-simulation-initialized MCMC
stationary_mcmc_intermediate <- function(n_iterations, N = 20, summary = "alpha") {
  initialization <- forward_intermediate(N)
  y <- initialization$y

  if (n_iterations == 0) {
    return(extract_summary_intermediate(initialization, summary))
  }

  current <- initialization
  logp_current <- joint_intermediate(current, y)

  for (i in seq_len(n_iterations)) {

    out <- mh_update(
      current, y, logp_current,
      function(s) { s$alpha <- s$alpha + rnorm(1, 0, 0.20); s }
    )
    current <- out$state; logp_current <- out$logp

    out <- mh_update(
      current, y, logp_current,
      function(s) { s$beta1 <- s$beta1 + rnorm(1, 0, 0.12); s }
    )
    current <- out$state; logp_current <- out$logp

    out <- mh_update(
      current, y, logp_current,
      function(s) { s$beta5 <- s$beta5 + rnorm(1, 0, 0.12); s }
    )
    current <- out$state; logp_current <- out$logp

    out <- mh_update(
      current, y, logp_current,
      function(s) { s$ell <- s$ell + rnorm(1, 0, 0.15); s }
    )
    current <- out$state; logp_current <- out$logp

    for (j in seq_along(current$z)) {
      out <- mh_update(
        current, y, logp_current,
        function(s) {
          s$z[j] <- s$z[j] + rnorm(1, 0, 0.30)
          s
        }
      )
      current <- out$state; logp_current <- out$logp
    }
  }

  extract_summary_intermediate(current, summary)
}
```

```{r}

# exact invariance test
exact_invariance_intermediate <- function(summary = "alpha", n_rep = 200, n_iter = 30, N = 20) {
  forward_only <- replicate(n_rep, stationary_mcmc_intermediate(0, N, summary))
  with_mcmc    <- replicate(n_rep, stationary_mcmc_intermediate(n_iter, N, summary))

  ks <- ks.test(forward_only, with_mcmc)

  data.frame(
    summary = summary,
    D_statistic = unname(ks$statistic),
    p_value = ks$p.value
  )
}

# run a few summaries
rbind(
  exact_invariance_intermediate("alpha"),
  exact_invariance_intermediate("beta1"),
  exact_invariance_intermediate("beta5"),
  exact_invariance_intermediate("sigma")
)

```

## 1.4 ESS

```{r}

# 1.6 ESS for intermediate model

ess_table_intermediate <- fit_hmc$summary(
  NULL,
  c("mean", "sd", "mcse_mean", "ess_bulk", "ess_tail", "rhat")
)

ess_table_intermediate[
  ess_table_intermediate$variable %in% c("alpha", "beta1", "beta5", "sigma"),
  c("variable", "mean", "sd", "mcse_mean", "ess_bulk", "ess_tail", "rhat")
]
```

## 3 Advanced Model

## 3.2 Model Implementation

```{r}

# -----------------------------
dat <- read_tsv("Data/Backup/backup.tsv", show_col_types = FALSE) %>%
  arrange(id, day, beep) %>%
  mutate(
    happy = as.integer(happy),
    row_index = row_number()
  )


# 1) Standardize exogenous predictors using observed values only

scale_to_numeric <- function(x) {
  m <- mean(x, na.rm = TRUE)
  s <- sd(x, na.rm = TRUE)
  (x - m) / s
}

dat <- dat %>%
  mutate(
    enthusiastic_std = scale_to_numeric(enthusiastic),
    pleased_std      = scale_to_numeric(pleased),
    relaxed_std      = scale_to_numeric(relaxed)
  )


# 2) Basic observed-y objects

N <- nrow(dat)

is_obs <- !is.na(dat$happy)
obs_index <- which(is_obs)
y_obs <- dat$happy[is_obs]
N_obs <- length(y_obs)


# 3) Build latent lag structure

has_lag1 <- rep(0L, N)
lag1_index <- rep(0L, N)

has_lag5 <- rep(0L, N)
lag5_index <- rep(0L, N)

for (i in seq_len(N)) {
  # lag 1: previous row must be same participant
  if (i > 1 && dat$id[i] == dat$id[i - 1]) {
    has_lag1[i] <- 1L
    lag1_index[i] <- i - 1L
  }

  # lag 5: five rows before must be same participant
  if (i > 5 && dat$id[i] == dat$id[i - 5]) {
    has_lag5[i] <- 1L
    lag5_index[i] <- i - 5L
  }
}


# 4) Build lagged exogenous predictors
#    Use previous-row values only if same participant and observed

enth_lag1 <- rep(0, N)
pleased_lag1 <- rep(0, N)
relaxed_lag1 <- rep(0, N)

has_enth_lag1 <- rep(0L, N)
has_pleased_lag1 <- rep(0L, N)
has_relaxed_lag1 <- rep(0L, N)

for (i in seq_len(N)) {
  if (i > 1 && dat$id[i] == dat$id[i - 1]) {
    if (!is.na(dat$enthusiastic_std[i - 1])) {
      enth_lag1[i] <- dat$enthusiastic_std[i - 1]
      has_enth_lag1[i] <- 1L
    }
    if (!is.na(dat$pleased_std[i - 1])) {
      pleased_lag1[i] <- dat$pleased_std[i - 1]
      has_pleased_lag1[i] <- 1L
    }
    if (!is.na(dat$relaxed_std[i - 1])) {
      relaxed_lag1[i] <- dat$relaxed_std[i - 1]
      has_relaxed_lag1[i] <- 1L
    }
  }
}


# 5) Final Stan data list

stan_data_model3 <- list(
  N = N,
  N_obs = N_obs,
  y_obs = y_obs,
  obs_index = obs_index,
  has_lag1 = has_lag1,
  lag1_index = lag1_index,
  has_lag5 = has_lag5,
  lag5_index = lag5_index,
  enth_lag1 = enth_lag1,
  pleased_lag1 = pleased_lag1,
  relaxed_lag1 = relaxed_lag1,
  has_enth_lag1 = has_enth_lag1,
  has_pleased_lag1 = has_pleased_lag1,
  has_relaxed_lag1 = has_relaxed_lag1
)


stopifnot(length(y_obs) == N_obs)
stopifnot(length(obs_index) == N_obs)
stopifnot(length(has_lag1) == N)
stopifnot(length(lag1_index) == N)
stopifnot(length(has_lag5) == N)
stopifnot(length(lag5_index) == N)
stopifnot(length(enth_lag1) == N)
stopifnot(length(pleased_lag1) == N)
stopifnot(length(relaxed_lag1) == N)
```

```{r}
#HMC
dir.create("stan_out", showWarnings = FALSE)

mod_model3 <- cmdstan_model("advanced_model.stan")

fit_hmc_model3 <- mod_model3$sample(
  data = stan_data_model3,
  chains = 2,
  seed = 405,
  iter_warmup =500,
  iter_sampling = 1000,
  refresh = 100,
  max_treedepth = 15,
  parallel_chains = 2,
  output_dir = "stan_out",
)
```

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(bayesplot)
  library(gridExtra)
  library(grid)
})

p1<-mcmc_trace(fit_hmc_model3$draws("gamma_enth")) + theme_minimal()
p1
```

```{r}
p2<-mcmc_rank_hist(fit_hmc_model3$draws("gamma_enth")) 
desktop <- file.path(Sys.getenv("USERPROFILE"), "Desktop")   # Windows

pdf(file.path(desktop, "mu_diagnostics.pdf"), width = 12, height = 5)
grid.arrange(p1, p2, ncol = 2)
dev.off()
```

```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(bayesplot)
  library(gridExtra)
  library(grid)
})

# save one diagnostic pdf for one parameter
save_diag_pdf <- function(fit_obj, param_name, file_prefix = "advanced") {
  p1 <- mcmc_trace(fit_obj$draws(param_name)) + theme_minimal()
  p2 <- mcmc_rank_hist(fit_obj$draws(param_name)) + theme_minimal()

  desktop <- file.path(Sys.getenv("USERPROFILE"), "Desktop")   # Windows
  out_file <- file.path(desktop, paste0(file_prefix, "_", param_name, "_diagnostics.pdf"))

  pdf(out_file, width = 12, height = 5)
  grid.arrange(p1, p2, ncol = 2)
  dev.off()

  cat("Saved:", out_file, "\n")
}

save_diag_pdf(fit_hmc_model3, "alpha")
save_diag_pdf(fit_hmc_model3, "beta1")
save_diag_pdf(fit_hmc_model3, "beta5")
save_diag_pdf(fit_hmc_model3, "gamma_enth")
save_diag_pdf(fit_hmc_model3, "gamma_pleased")
save_diag_pdf(fit_hmc_model3, "gamma_relaxed")
save_diag_pdf(fit_hmc_model3, "sigma")
```

Exact Invar Test

```{r}
set.seed(405)

# build a small debug version with N = 30

N_small <- 20

stan_data_ext_small <- stan_data_model3

keep_obs <- which(stan_data_model3$obs_index <= N_small)

stan_data_ext_small$N <- N_small
stan_data_ext_small$N_obs <- length(keep_obs)
stan_data_ext_small$y_obs <- stan_data_model3$y_obs[keep_obs]
stan_data_ext_small$obs_index <- stan_data_model3$obs_index[keep_obs]

stan_data_ext_small$has_lag1 <- stan_data_model3$has_lag1[1:N_small]
stan_data_ext_small$lag1_index <- stan_data_model3$lag1_index[1:N_small]

stan_data_ext_small$has_lag5 <- stan_data_model3$has_lag5[1:N_small]
stan_data_ext_small$lag5_index <- stan_data_model3$lag5_index[1:N_small]

stan_data_ext_small$has_enth_lag1 <- stan_data_model3$has_enth_lag1[1:N_small]
stan_data_ext_small$enth_lag1 <- stan_data_model3$enth_lag1[1:N_small]

stan_data_ext_small$has_pleased_lag1 <- stan_data_model3$has_pleased_lag1[1:N_small]
stan_data_ext_small$pleased_lag1 <- stan_data_model3$pleased_lag1[1:N_small]

stan_data_ext_small$has_relaxed_lag1 <- stan_data_model3$has_relaxed_lag1[1:N_small]
stan_data_ext_small$relaxed_lag1 <- stan_data_model3$relaxed_lag1[1:N_small]


# fixed cutpoints

cuts_fixed <- c(-2.5, -1.5, -0.5, 0.5, 1.5, 2.5)


# ordered logistic helpers

ordered_logistic_probs <- function(eta, cuts = cuts_fixed) {
  p <- numeric(length(cuts) + 1)
  p[1] <- plogis(cuts[1] - eta)
  for (k in 2:length(cuts)) {
    p[k] <- plogis(cuts[k] - eta) - plogis(cuts[k - 1] - eta)
  }
  p[length(cuts) + 1] <- 1 - plogis(cuts[length(cuts)] - eta)
  pmax(p, 1e-12)
}

rordered_logistic1 <- function(eta, cuts = cuts_fixed) {
  sample.int(length(cuts) + 1, 1, prob = ordered_logistic_probs(eta, cuts))
}

log_ordered_logistic_vec <- function(y, eta, cuts = cuts_fixed) {
  probs <- mapply(function(yy, ee) ordered_logistic_probs(ee, cuts)[yy], y, eta)
  sum(log(pmax(probs, 1e-12)))
}


# reconstruct transformed params and latent path
# matching advanced_model.stan

make_x_extended <- function(state, dat) {
  alpha <- state$alpha_raw
  beta1 <- 0.45 * tanh(state$beta1_raw)
  beta5 <- 0.45 * tanh(state$beta5_raw)
  sigma <- exp(state$ell)

  N <- dat$N
  x <- numeric(N)
  mean_x <- numeric(N)

  for (i in seq_len(N)) {
    mean_x[i] <- alpha

    if (dat$has_lag1[i] == 1) {
      mean_x[i] <- mean_x[i] + beta1 * x[dat$lag1_index[i]]
    }

    if (dat$has_lag5[i] == 1) {
      mean_x[i] <- mean_x[i] + beta5 * x[dat$lag5_index[i]]
    }

    if (dat$has_enth_lag1[i] == 1) {
      mean_x[i] <- mean_x[i] + state$gamma_enth * dat$enth_lag1[i]
    }

    if (dat$has_pleased_lag1[i] == 1) {
      mean_x[i] <- mean_x[i] + state$gamma_pleased * dat$pleased_lag1[i]
    }

    if (dat$has_relaxed_lag1[i] == 1) {
      mean_x[i] <- mean_x[i] + state$gamma_relaxed * dat$relaxed_lag1[i]
    }

    x[i] <- mean_x[i] + sigma * state$z[i]

    if (!is.finite(x[i])) return(NULL)
  }

  list(
    alpha = alpha,
    beta1 = beta1,
    beta5 = beta5,
    sigma = sigma,
    mean_x = mean_x,
    x = x
  )
}


# forward simulator

forward_extended <- function(dat, cuts = cuts_fixed) {
  repeat {
    state <- list(
      alpha_raw = rnorm(1, 0, 1),
      beta1_raw = rnorm(1, 0, 1),
      beta5_raw = rnorm(1, 0, 1),
      gamma_enth = rnorm(1, 0, 0.5),
      gamma_pleased = rnorm(1, 0, 0.5),
      gamma_relaxed = rnorm(1, 0, 0.5),
      ell = log(rexp(1, rate = 2)),
      z = rnorm(dat$N)
    )

    comp <- make_x_extended(state, dat)

    if (!is.null(comp)) {
      y_obs <- sapply(comp$x[dat$obs_index], rordered_logistic1, cuts = cuts)
      state$y_obs <- y_obs
      return(state)
    }
  }
}

# # joint density

joint_extended <- function(state, y_obs, dat, cuts = cuts_fixed) {
  comp <- make_x_extended(state, dat)
  if (is.null(comp)) return(-Inf)

  dnorm(state$alpha_raw, 0, 1, log = TRUE) +
    dnorm(state$beta1_raw, 0, 1, log = TRUE) +
    dnorm(state$beta5_raw, 0, 1, log = TRUE) +
    dnorm(state$gamma_enth, 0, 0.5, log = TRUE) +
    dnorm(state$gamma_pleased, 0, 0.5, log = TRUE) +
    dnorm(state$gamma_relaxed, 0, 0.5, log = TRUE) +
    dexp(comp$sigma, rate = 2, log = TRUE) +
    state$ell +   # Jacobian for sigma = exp(ell)
    sum(dnorm(state$z, 0, 1, log = TRUE)) +
    log_ordered_logistic_vec(y_obs, comp$x[dat$obs_index], cuts)
}


# summary extractor

extract_summary_extended <- function(
    state, dat,
    summary = c("alpha", "beta1", "beta5",
                "gamma_enth", "gamma_pleased", "gamma_relaxed",
                "sigma", "xbar", "x6")
) {
  summary <- match.arg(summary)
  comp <- make_x_extended(state, dat)

  switch(
    summary,
    alpha = comp$alpha,
    beta1 = comp$beta1,
    beta5 = comp$beta5,
    gamma_enth = state$gamma_enth,
    gamma_pleased = state$gamma_pleased,
    gamma_relaxed = state$gamma_relaxed,
    sigma = comp$sigma,
    xbar = mean(comp$x),
    x6 = comp$x[min(6, length(comp$x))]
  )
}


# one MH update helper

mh_update <- function(state, y_obs, logp_current, proposal_fun, dat) {
  proposal <- proposal_fun(state)
  logp_prop <- joint_extended(proposal, y_obs, dat)

  if (log(runif(1)) < (logp_prop - logp_current)) {
    list(state = proposal, logp = logp_prop)
  } else {
    list(state = state, logp = logp_current)
  }
}

# forward-simulation-initialized MCMC

stationary_mcmc_extended <- function(n_iterations, dat = stan_data_ext_small, summary = "alpha") {
  initialization <- forward_extended(dat)
  y_obs <- initialization$y_obs

  if (n_iterations == 0) {
    return(extract_summary_extended(initialization, dat, summary))
  }

  current <- initialization
  logp_current <- joint_extended(current, y_obs, dat)

  for (iter in seq_len(n_iterations)) {

    out <- mh_update(
      current, y_obs, logp_current,
      function(s) { s$alpha_raw <- s$alpha_raw + rnorm(1, 0, 0.15); s },
      dat
    )
    current <- out$state; logp_current <- out$logp

    out <- mh_update(
      current, y_obs, logp_current,
      function(s) { s$beta1_raw <- s$beta1_raw + rnorm(1, 0, 0.12); s },
      dat
    )
    current <- out$state; logp_current <- out$logp

    out <- mh_update(
      current, y_obs, logp_current,
      function(s) { s$beta5_raw <- s$beta5_raw + rnorm(1, 0, 0.12); s },
      dat
    )
    current <- out$state; logp_current <- out$logp

    out <- mh_update(
      current, y_obs, logp_current,
      function(s) { s$gamma_enth <- s$gamma_enth + rnorm(1, 0, 0.08); s },
      dat
    )
    current <- out$state; logp_current <- out$logp

    out <- mh_update(
      current, y_obs, logp_current,
      function(s) { s$gamma_pleased <- s$gamma_pleased + rnorm(1, 0, 0.08); s },
      dat
    )
    current <- out$state; logp_current <- out$logp

    out <- mh_update(
      current, y_obs, logp_current,
      function(s) { s$gamma_relaxed <- s$gamma_relaxed + rnorm(1, 0, 0.08); s },
      dat
    )
    current <- out$state; logp_current <- out$logp

    out <- mh_update(
      current, y_obs, logp_current,
      function(s) { s$ell <- s$ell + rnorm(1, 0, 0.10); s },
      dat
    )
    current <- out$state; logp_current <- out$logp

    for (j in seq_along(current$z)) {
      out <- mh_update(
        current, y_obs, logp_current,
        function(s) {
          s$z[j] <- s$z[j] + rnorm(1, 0, 0.25)
          s
        },
        dat
      )
      current <- out$state
      logp_current <- out$logp
    }
  }

  extract_summary_extended(current, dat, summary)
}


# exact invariance test
# same logic as before

exact_invariance_extended <- function(summary = "alpha",
                                      n_rep = 200,
                                      n_iter = 30,
                                      dat = stan_data_ext_small) {
  forward_only <- replicate(
    n_rep,
    stationary_mcmc_extended(0, dat, summary)
  )

  with_mcmc <- replicate(
    n_rep,
    stationary_mcmc_extended(n_iter, dat, summary)
  )

  ks <- ks.test(forward_only, with_mcmc)

  data.frame(
    summary = summary,
    D_statistic = unname(ks$statistic),
    p_value = ks$p.value
  )
}


# run the test
# consistent with earlier scale: N = 30

rbind(
  exact_invariance_extended("alpha", n_rep = 200, n_iter = 30),
  exact_invariance_extended("beta1", n_rep = 200, n_iter = 30),
  exact_invariance_extended("beta5", n_rep = 200, n_iter = 30),
  exact_invariance_extended("gamma_enth", n_rep = 200, n_iter = 30),
  exact_invariance_extended("gamma_pleased", n_rep = 200, n_iter = 30),
  exact_invariance_extended("gamma_relaxed", n_rep = 200, n_iter = 30),
  exact_invariance_extended("sigma", n_rep = 200, n_iter = 30)
)


```

ESS

```{r}

# ESS for extended model

ess_table_extended <- fit_hmc_model3$summary(
  NULL,
  c("mean", "sd", "mcse_mean", "ess_bulk", "ess_tail", "rhat")
)

ess_table_extended[
  ess_table_extended$variable %in% c(
    "alpha", "beta1", "beta5",
    "gamma_enth", "gamma_pleased", "gamma_relaxed",
    "sigma"
  ),
  c("variable", "mean", "sd", "mcse_mean", "ess_bulk", "ess_tail", "rhat")
]
```