\documentclass[12pt,letterpaper,fleqn]{article}

% Page layout
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

% Math packages
\usepackage{amsmath,amssymb}

% Figures and tables
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{subcaption}

% Citations
\usepackage[american]{babel}
\usepackage{csquotes}
\usepackage[
style=apa,
backend=biber,
sorting=nyt
]{biblatex}
\addbibresource{references.bib}

% Hyperlinks
\usepackage{hyperref}

% Left aligned equations (fleqn)
\setlength{\mathindent}{0pt}

% Custom commands
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\Corr}{\mathrm{Corr}}
\newcommand{\Prb}{\mathrm{Pr}}

%------------------------------------------------
% Title block
%------------------------------------------------

\title{\large\bfseries
hhh
}

\author{
\small
Marc Shi,  Yu Chang \\
 \small
University of British Columbia
}

\date{\small
STAT 405: Bayesian Statistics \\
Instructor: Dr.~Alexandre Bouchard-Côté \\
April 2026
}

%------------------------------------------------

\begin{document}

\maketitle

\section{Literature Review}

Emotions play a fundamental role in psychological well-being and daily functioning. Understanding how emotional states evolve over time has become an important topic in psychology and behavioral science. Traditionally, research on affect has focused on static summaries of emotion, such as average levels of positive or negative affect measured through questionnaires. However, these approaches ignore the temporal dynamics of emotional experiences. In reality, emotional states fluctuate continuously and are influenced by both internal processes and external factors. As a result, recent work has increasingly focused on modeling emotional experiences as dynamic systems, where emotions at one moment may influence emotions at later moments.

One important line of research investigates emotion networks, which conceptualize emotions as nodes connected through dynamic relationships that evolve over time. Kullar et al. (2024) used an ESM design to study within-person emotion dynamics over a two-week period. Participants received five prompts per day via smartphone and rated the intensity of several emotions, including enthusiastic, happy, pleased, relaxed, etc. These repeated measurements produced individual-level time series of emotional experiences. To model the interactions among these emotions, the authors applied the Group Iterative Multiple Model Estimation (GIMME) framework, which runs auto model selection. In this approach, each emotion variable is treated as an observed node in a dynamic network, and both lagged (temporal) and contemporaneous relationships between emotions are estimated. This modeling strategy allows the analysis to capture how emotional states influence one another across time as well as within the same time point, while also allowing for individual differences in these dynamic emotion networks. The results suggest that emotion dynamics vary substantially across individuals and may form patterns that do not align neatly with traditional diagnostic categories.

Despite its strength, this modeling approahc implicitly assume that the observed emotion scores directly represent the true random emotional state of an individual. From a statistical perspective, this assumption can be viewd as naive because psyuchological measuemenets are often subject to measurement error. Self-reported emotion rating may reflect ot only the true emotional state but also random reporting noise. As a result, a natural Bayesian alternate is to introduce a latent-state formulation, where the true emotional state is treated as unobserved variable and the reported emotion scores are viewd as noisy observation of that latent state with a likelihood distribution conditional on the underlying eomtional state. This perspective allows 



\section{Exploratory Data Analysis}

Describe the dataset and show preliminary plots.



\section{Methodology}

Describe the models used in the analysis.

\section{Results}

Present forecasting results and model comparisons.

\section{Model Diagnostics}

Discuss stationarity, ACF/PACF, and residual diagnostics.

\section{Conclusion}

Summarize findings and discuss implications.

\section*{Author Contributions}

Describe contributions of each team member.

\printbibliography

\end{document}